---
author: Aymen Ben Brik 
date: "`r Sys.Date()`"
output: html_document
title: Time Series Analysis of Atlanta Temperatures
---

# Introduction

This document analyses monthly temperature data from Atlanta.\
The objective is to:

-   Import and format the data
-   Create a time series object
-   Estimate the **trend**
-   Estimate the **seasonality**
-   Remove trend and seasonality
-   Study the **autocorrelation of residuals**

------------------------------------------------------------------------

# 1. Data Import and Preparation

``` {r}
# Read the text file containing Atlanta average temperatures
data = read.table("AvTempAtlanta.txt", header = T)

# Display column names
names(data)

# Visualize the dataset
View(data)

# Remove column 1 and 14 (often year and annual average)
data = data[,-c(1,14)]

# Check the data after removing columns
View(data)

# Show help for as.vector
help(as.vector)

# Transpose the matrix and convert it into a vector
# This ensures observations are ordered month by month
data = as.vector(t(data))

# View the resulting vector
View(data)
```

------------------------------------------------------------------------

# 2. Create the Time Series

``` {r}
# Create a monthly time series starting in 1879
temp = ts(data, start = 1879, frequency = 12)

# Verify object class
class(temp)

# Plot the time series
ts.plot(temp, ylab = "temperature")
```

------------------------------------------------------------------------

# 3. Trend Estimation

## Create Time Points

``` {r}
# Create equally spaced time points
time.pts = c(1:length(temp))

# Normalize time between 0 and 1 for numerical stability
time.pts = (time.pts - min(time.pts))/max(time.pts)

# Display time points
time.pts
```

------------------------------------------------------------------------

## 3.1 Moving Average Trend

``` {r}
# Help for kernel smoothing
help(ksmooth)

# Apply kernel smoothing (box kernel ≈ moving average)
mav.fit = ksmooth(time.pts, temp, kernel = "box")

# Convert fitted values into a time series
temp.fit.mav = ts(mav.fit$y, start = 1879, frequency = 12)
```

### Visualize Trend

``` {r}
# Plot original series
ts.plot(temp, ylab = "temperature")

# Add moving average trend
lines(temp.fit.mav, lwd = 2, col = "purple")

# Horizontal reference line
abline(temp.fit.mav[1], 0, lwd = 2, col = "blue")
```

------------------------------------------------------------------------

## 3.2 Parametric Quadratic Trend

``` {r}
# Linear and quadratic time variables
x1 = time.pts
x2 = time.pts^2

# Fit quadratic regression model
lm.fit = lm(temp ~ x1 + x2)

# Model summary
summary(lm.fit)
```

### Plot Trend

``` {r}
# Extract fitted values
temp.fit.lm = ts(fitted(lm.fit), start = 1879, frequency = 12)

# Plot original series
ts.plot(temp, ylab = "temperature")

# Add quadratic trend
lines(temp.fit.lm, lwd = 2, col = "green")

# Reference line
abline(temp.fit.lm[1], 0, lwd = 2, col = "blue")
```

------------------------------------------------------------------------

## 3.3 Local Polynomial Trend (LOESS)

``` {r}
# Local regression trend estimation
loc.fit = loess(temp ~ time.pts)

# Extract fitted values
temp.loc.fit = ts(fitted(loc.fit), start = 1879, frequency = 12)
```

------------------------------------------------------------------------

## 3.4 Spline Trend Estimation (GAM)

``` {r}
# Load mgcv package
library(mgcv)

# Fit GAM model with spline trend
gam.fit = gam(temp ~ s(time.pts))

# Extract fitted values
temp.fit.gam = ts(fitted(gam.fit), start = 1879, frequency = 12)
```

### Plot Nonparametric Trends

``` {r}
ts.plot(temp, ylab = "temperature")

# LOESS trend
lines(temp.loc.fit, lwd = 2, col = "red")

# Spline trend
lines(temp.fit.gam, lwd = 2, col = "yellow")

# Reference line
abline(temp.loc.fit[1], 0, lwd = 2, col = "blue")
```

------------------------------------------------------------------------

# 4. Compare Trend Estimates

``` {r}
# Combine trend values to define plot limits
all.val = c(temp.fit.mav, temp.fit.mav, temp.loc.fit, temp.fit.gam)

ylim = c(min(all.val), max(all.val))

# Plot moving average trend
ts.plot(temp.fit.mav, lwd = 2, ylim = ylim, ylab = "temperature", col = "blue")

# Quadratic trend
lines(temp.fit.lm, lwd = 2, col = "green")

# LOESS trend
lines(temp.loc.fit, lwd = 2, col = "red")

# Spline trend
lines(temp.fit.gam, lwd = 2, col = "yellow")
```

------------------------------------------------------------------------

# 5. Seasonality Estimation

``` {r}
library(TSA)

# Identify the month of each observation
month = season(temp)
```

## Seasonal Means Model

``` {r}
# Linear model with monthly dummies
model1 = lm(temp ~ month)
summary(model1)

# Model without intercept (direct monthly means)
model2 = lm(temp ~ month - 1)
summary(model2)
```

------------------------------------------------------------------------

## Harmonic Seasonal Model

``` {r}
# First harmonic (sin/cos)
har1 = harmonic(temp, 1)

model3 = lm(temp ~ har1)
summary(model3)

# Second harmonic
har2 = harmonic(temp, 2)

model4 = lm(temp ~ har2)
summary(model4)
```

------------------------------------------------------------------------

## Compare Seasonal Models

``` {r}
# Seasonal means coefficients
st1 = coef(model2)

# Harmonic seasonal pattern
st2 = fitted(model4)[1:12]

plot(1:12, st1, lwd = 2, type = "l", xlab = "month", ylab = "Seasonality")
lines(1:12, st2, lwd = 2, col = "blue")
```

------------------------------------------------------------------------

# 6. Trend + Seasonality Model

``` {r}
# Parametric trend + harmonic seasonality
lm.fit2 = lm(temp ~ x1 + x2 + har2)

summary(lm.fit2)

# Residual process
dif.fit.lm = ts(temp - fitted(lm.fit2), start = 1879, frequency = 12)

ts.plot(dif.fit.lm, ylab = "Residual Process")
```

------------------------------------------------------------------------

## Nonparametric Trend + Seasonality

``` {r}
gam.fit2 = gam(temp ~ s(time.pts) + har2)

dif.gam.fit2 = ts(temp - fitted(gam.fit2), start = 1879, frequency = 12)

#lines(dif.gam.fit2, col = "blue")
```

------------------------------------------------------------------------

# 7. Autocorrelation Analysis

``` {r}
par(mfrow = c(3,1))

# ACF of original series
acf(temp, lag.max = 4*12, main = "")

# ACF of residuals (parametric model)
acf(dif.fit.lm, lag.max = 4*12)

# ACF of residuals (GAM model)
acf(dif.gam.fit2, lag.max = 4*12)
```
